\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-06-23 paper.qmd
fig 3 re-organised

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(quollr)
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)
library(tidyr)

library(Rtsne)
library(uwot)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)
library(grid)
```

@@

hex_grid_with_counts <- 
  dplyr::left_join(hex_grid,
                   counts_df, 
                   by = c("hex_poly_id" = "hb_id"))

@@

```{r}
#| label: fig-NLDR-scurve
#| echo: false
#| fig-cap: "Key steps for constructing the model on the UMAP layout ($k=2$) of the S-curve data: (a) data, (b) hexagon bins, (c) bin centroids, and (d) triangulated centroids."
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%

@@

It is beneficial to define the algorithm on data having a standard scale. Here the variables are scaled to [0, 1], but the upper bound can incorporate the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$, by setting them to $(y_{1,\text{max}}, y_{2,\text{max}}, ..., y_{k,\text{max}})$. When $k=2$ which is assumed for hexagon binning, $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = \frac{r_2}{r_1}$, as observed in @fig-NLDR-scurve. <!-- \times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) The scaling of data should take the size of the hexagons into account, but choice of number of bins should. -->

@@

```{r}
# Code to draw illustration for notation
hex_param_vis <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
    geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
    geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "#31a354", size = 0.9) +
    geom_point(data = all_centroids_df |> dplyr::filter(hexID == 1), aes(x = c_x, y = c_y), color = "black") + ## starting point
    geom_rect(aes(xmin = 0 - 0.03, xmax = 1 - 0.03, ymin = 0 - 0.05, ymax = r2- 0.05), fill = "white", color = "black", alpha = 0, linewidth = 0.7) +
    geom_segment(x = 0 - 0.03, y = -0.35, xend = 1 - 0.03, yend = -0.35,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+
    annotate("text", x=0.5, y=-0.45, label= expression(r[1]), color = "black") +
    geom_segment(x = -0.25, y = 0 - 0.05, xend = -0.25, yend = r2 - 0.05,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ 
    annotate("text", x=-0.35, y=1, label= expression(r[2]), color = "black") +
    geom_segment(x = -0.1 + 0.2087578, y = -0.15, xend = -0.1 + 0.2087578*2, yend = -0.15,
                   arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a1 = 0.2087578
    annotate("text", x=0.2, y=-0.25, label= expression(a[1]), color = "black") +
    geom_segment(x = -0.15, y = -0.1*r2 + 0.1807896*2, xend = -0.15, yend = -0.1*r2 + 0.1807896*3,
                     arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a2 = 0.1807896
    annotate("text", x=-0.15, y=0.45, label= expression(a[2]), color = "black") +
    annotate("text", x=-0.18, y=-0.25, label= expression(group("(", list(s[1], s[2]), ")")), color = "black") ## starting point
```

```{r}
#| label: fig-hex-param
#| fig-cap: "Illustration of the components of the hexagon grid with notation."
#| out-height: 30%
#| fig-width: 3
#| fig-height: 4.5
#| fig-pos: H

>>>>

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(quollr)
library(tibble)
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)
library(tidyr)

library(Rtsne)
library(uwot)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)
library(grid)
library(conflicted)
conflicts_prefer(dplyr::filter)
```

@@

hex_grid_with_counts <- 
  left_join(hex_grid,
            counts_df, 
            by = c("hex_poly_id" = "hb_id"))

@@

```{r}
#| label: fig-NLDR-scurve
#| echo: false
#| fig-cap: "Key steps for constructing the model on the UMAP layout ($k=2$): (a) data, (b) hexagon bins, (c) bin centroids, and (d) triangulated centroids. The S-curve data is shown."
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%

@@

Because we are working with distances between points, starting with data having a standard scale, e.g. [0, 1], is recommended. When the aspect ratio of the NLDR $(r_1, r_2, ..., r_k)$ is meaningful, the data can be scaled to take this into account. When $k=2$, as in hexagon binning, the default range is $[0, y_{i,\text{max}}], i=1,2$, where $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = \frac{r_2}{r_1}$ (@fig-NLDR-scurve). <!-- \times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) The scaling of data should take the size of the hexagons into account, but choice of number of bins should. -->

@@

```{r}
# Code to draw illustration for notation
start_pt <- all_centroids_df |> 
  filter(hexID == 1)
d_rect <- tibble(x1min = 0, 
                 x1max = 1,
                 x2min = 0,
                 x2max = r2)
c_start <- tibble(s1 = 0.03, s2 = 0.05)
a1 <- tibble(x = all_centroids_df$c_x[21],
             xend = all_centroids_df$c_x[22],
             y = all_centroids_df$c_y[21],
             yend = all_centroids_df$c_y[21],
             label = expression(a[1]))
a2 <- tibble(x = all_centroids_df$c_x[27],
             xend = all_centroids_df$c_x[27],
             y = all_centroids_df$c_y[27],
             yend = all_centroids_df$c_y[33],
             label = expression(a[2]))
hex_param_vis <- ggplot() + 
    geom_polygon(data = hex_grid, 
                        aes(x = x, 
                            y = y, 
                            group = hex_poly_id),
                 fill = "white", 
                 color = "#bdbdbd") +
    geom_point(data = all_centroids_df, aes(
      x = c_x, 
      y = c_y), 
      color = "#31a354", size = 0.9) +
    geom_point(data = start_pt, aes(x = c_x, 
                                    y = c_y), 
               color = "black") + 
    geom_rect(data=d_rect, 
              aes(xmin = x1min - c_start$s1,# - c_start$s1, 
                  xmax = x1max - c_start$s1,# - c_start$s1, 
                  ymin = x2min - c_start$s2,# - c_start$s2, 
                  ymax = x2max - c_start$s2),# - c_start$s2), 
              fill = "white", 
              color = "black", 
              alpha = 0, 
              linewidth = 0.7) +
    geom_point(data=d_rect, aes(x=x1min - c_start$s1, 
                                y=x2min - c_start$s2)) + 
    geom_point(data=d_rect, aes(x=x1max - c_start$s1, 
                                y=x2min - c_start$s2)) + 
    geom_point(data=d_rect, aes(x=x1min - c_start$s1, 
                                y=x2max - c_start$s2)) + 
    annotate("text", x=d_rect$x1min - c_start$s1, 
                     y=d_rect$x2min - c_start$s2,
                     label = "(0,0)", 
             hjust=-0.1, vjust=-0.3) + 
    annotate("text", x=d_rect$x1max - c_start$s1, 
                     y=d_rect$x2min - c_start$s2,
                     label = "(0,1)", 
             hjust=1.1, vjust=-0.3) + 
    annotate("text", x=d_rect$x1min - c_start$s1, 
                     y=d_rect$x2max - c_start$s2,
                     label = expression(group("(", 
                        list(0, y[2][max]),")")), 
            hjust=-0.1, vjust=1.2) + 
    geom_segment(data=d_rect, aes(
      x = x1min  - c_start$s1, # 0 - 0.03, 
      y = -0.35, 
      xend = x1max - c_start$s1, #1 - 0.03, 
      yend = -0.35), #-0.35),
      arrow = arrow(length = unit(0.03, "npc"),
                               ends = "both"), 
                 color = "black")+
    annotate("text", x=0.5, y=-0.45, 
             label = expression(r[1]), color = "black") +
    geom_segment(data=d_rect, aes(
      x = -0.25, 
      y = x2min - c_start$s2, #0 - 0.05, 
      xend = -0.25, 
      yend = x2max - c_start$s2), #r2 - 0.05),
      arrow = arrow(length = unit(0.03, "npc"),
                       ends = "both"), 
                 color = "black")+ 
    annotate("text", x=-0.35, y=1, 
             label = expression(r[2]), color = "black") +
    geom_segment(data = a1, aes(
      x = x, #-0.1 + 0.2087578, 
      y = y, #-0.15, 
      xend = xend, #-0.1 + 0.2087578*2, 
      yend = yend), #-0.15),
      arrow = arrow(length = unit(0.03, "npc"),
        ends = "both"), 
        color = "black")+ # a1 = 0.2087578
    annotate("text", 
             x=(a1$x+a1$xend)/2, 
             y=a1$y, 
             label = expression(a[1]), 
             color = "black",
             vjust = 1.2) +
    geom_segment(data = a2, aes(
      x = x, #-0.15, 
      y = y, #-0.1*r2 + 0.1807896*2, 
      xend = xend, #-0.15, 
      yend = yend), #-0.1*r2 + 0.1807896*3),
      arrow = arrow(length = unit(0.03, "npc"),
                               ends = "both"), 
      color = "black") + # a2 = 0.1807896
    annotate("text", x=a2$x, y=(a2$y+a2$yend)/2, 
             label = expression(a[2]), 
             color = "black", hjust=-0.2) +
    annotate("text", x=-0.18, y=-0.25, 
      label = expression(group("(", list(s[1], s[2]), ")")),
      color = "black") 

```{r}
#| label: fig-hex-param
#| fig-cap: "The components of the hexagon grid illustrating notation."
#| out-height: 30%
#| fig-width: 2.5
#| fig-height: 3.75
#| fig-pos: H

### 2024-06-23 paper.tex
fig 3 re-organised

\caption{\label{fig-NLDR-scurve}Key steps for constructing the model on
the UMAP layout (\(k=2\)) of the S-curve data: (a) data, (b) hexagon
bins, (c) bin centroids, and (d) triangulated centroids.}

@@

It is beneficial to define the algorithm on data having a standard
scale. Here the variables are scaled to {[}0, 1{]}, but the upper bound
can incorporate the aspect ratio produced by the NLDR
\((r_1, r_2, ..., r_k)\), by setting them to
\((y_{1,\text{max}}, y_{2,\text{max}}, ..., y_{k,\text{max}})\). When
\(k=2\) which is assumed for hexagon binning, \(y_{1,\text{max}}=1\) and
\(y_{2,\text{max}} = \frac{r_2}{r_1}\), as observed in
Figure~\ref{fig-NLDR-scurve}.

@@

\caption{\label{fig-hex-param}Illustration of the components of the
hexagon grid with notation.}

>>>>

\caption{\label{fig-NLDR-scurve}Key steps for constructing the model on
the UMAP layout (\(k=2\)): (a) data, (b) hexagon bins, (c) bin
centroids, and (d) triangulated centroids. The S-curve data is shown.}

@@

Because we are working with distances between points, starting with data
having a standard scale, e.g.~{[}0, 1{]}, is recommended. When the
aspect ratio of the NLDR \((r_1, r_2, ..., r_k)\) is meaningful, the
data can be scaled to take this into account. When \(k=2\), as in
hexagon binning, the default range is \([0, y_{i,\text{max}}], i=1,2\),
where \(y_{1,\text{max}}=1\) and \(y_{2,\text{max}} = \frac{r_2}{r_1}\)
(Figure~\ref{fig-NLDR-scurve}).

@@

\caption{\label{fig-hex-param}The components of the hexagon grid
illustrating notation.}
